Gender disparity in computational biology research publications

### 2017-03-09 manuscript.tex
fix latex tables

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}

@@

\begin{document}
\maketitle
\section{Abstract}
While women are generally underrepresented in STEM fields, there are noticeable differences between fields. For instance, the gender ratio in biology is more balanced than in computer science. We were interested in how this difference is reflected in the interdisciplinary field of computational/quantitative biology. To this end, we examined the proportion of female authors in publications from the PubMed and arXiv databases. There are fewer female authors on research papers in computational biology, as compared to biology in general. This is true across authorship position, year, and journal impact factor. A comparison with arXiv shows that quantitative biology papers have a higher ratio of female authors than computer science papers, placing computational biology in between its two parent fields in terms of gender representation. Both in biology and in computational biology, a female last author increases the probability of other authors on the paper being female, pointing to a potential role of female PIs in influencing the gender balance.

\section{Introduction}
There is ample literature on the underrepresentation of women in STEM fields and the biases contributing to it. Those biases, though often subtle, are pervasive in several ways: they are often held and perpetuated by both men and women, and they are apparent across all aspects of academic and scientific practice. Undergraduate students show bias in favor of men both when rating their peers [1] and their professors [2]. Professors, in turn, are more likely to respond to e-mail from prospective students who are male [3]. They also show gender bias when hiring staff and deciding on a starting salary [4].

@@

\section{Results and Discussion}
In order to determine if there is a difference in the gender of authors in computational biology compared to biology as a whole, we used data from Pubmed, a database of biology and biomedical publications administered by the US National Library of Medicine. Pubmed uses Medical Subject Heading (MeSH) terms to classify individual papers by subject. The MeSH term “Computational Biology” is a subset of “Biology” and was introduced in 1997, so we restricted our analysis to primary articles published after this date (see Supplementary Figure 1A-B, Materials and Methods).

@@

Table 1

One possible explanation for the difference in male and female authorship position might be a difference in role models or mentors. If true, we would expect studies with a female principal investigator to be more likely to attract female collaborators. Conventionally in biology,  the last author on a publication is the principal investigator on the project. Therefore, we looked at two subsets of our data: publications with a female last author ($P_{female}$ > 0.8) and those with a male last author ($P_{female}$ < 0.2). We found that women were substantially more likely to be authors at every other position if the paper had a female last author than if the last author was male (Figure 1C, Table 2). It is possible that female trainees are be more likely to pursue computational biology if they have a mentor that is also female. Since women are less likely to be senior authors, this might reduce the proportion of women overall.

Table 2

@@

Table 3

@@

Since biology attracts more women than computer science, we suspect that many women initially decide to study biology and later become interested in computational biology. If this is the case, understanding what factors influence the field of study will provide useful insight when designing interventions to help narrow the gender gap in computer science and computational biology.
Materials and Methods
Datasets
Biology publications 1997-2014 (bio)
This dataset [13] contains all English language publications under the MeSH term "Biology" published between 1997 and 2014, excluding many non-primary sources. Downloaded 12 February, 2016. Search term: ("Biology"[Mesh]) NOT (Review[ptyp] OR Comment[ptyp] OR Editorial[ptyp] OR Letter[ptyp] OR Case Reports[ptyp] OR News[ptyp] OR "Biography" [Publication Type]) AND ("1997/01/01"[PDAT] : "2014/12/31"[PDAT]) AND english[language]

Computational biology publications 1997-2014 (comp)
Same as above [13], except using MeSH term "Computational Biology". Only uses papers where this is is a major term. Date range was selected because this MeSH term was introduced in 1997. Downloaded 12 February, 2016. Search term: ("Computational Biology"[Majr]) NOT (Review[ptyp] OR Comment[ptyp] OR Editorial[ptyp] OR Letter[ptyp] OR Case Reports[ptyp] OR News[ptyp] OR "Biography" [Publication Type]) AND ("1997/01/01"[PDAT] : "2014/12/31"[PDAT]) AND english[language]

Medical Papers
Subset of author and gender data from Filardo et.al [12]. This dataset did not contain author first names or unique publication identifiers. We searched pubmed for the title, author and publication date, and were able to identify 2155/3153 publications to analyze. Publications with no matching search results or with multiple matching search results were excluded.

arXiv Quantitative Biology (q-bio)
This dataset [14] contains all preprints with the label “q-bio” from 2003 (when the section was introduced) to 2014. Downloaded on 10 June, 2016.

arXiv CS (cs)
This dataset [14] contains all preprints with the label “cs” from 2003 to 2014. Downloaded on 10 June, 2016.
Gender Inference
Genders were determined using Gender-API (http://gender-api.com), which compares first names to a database compiled from government sources as well as from crawling social media profiles and returns a gender probability and a measure of confidence based on the number of times the name appears in the database. The API was queried with the 74,760 unique first names in the dataset (24 May, 2016).

@@

Figures
Figure 1
Mean probability that an author in a given position is female for primary articles indexed in Pubmed with the MeSH term Biology (black) or Computational Biology (grey). The bio dataset is inclusive of papers in the comp dataset. Error bars represent 95\% confidence intervals.
Mean probability that an author is female for publications in a given year. Error bars represent 95\% confidence intervals.
Mean probability that the first (F), second (S), penultimate (P) or other (O) author is female for publications where the last author is male ($P_{female}$ < 0.2) or female ($P_{female}$ > 0.8). Papers where the gender of the last author was uncertain or could not be determined were excluded. Error bars represent 95\% confidence intervals.
Figure 2
Mean probability that an author is female for every journal that had at least 1000 authors in our dataset. Grey bars represent journals that have the words “Bioinformatics,” “Computational,” “Computer,” “System(s),” or “-omic(s)” in their title. Vertical line represents the median for female author representation. See also S1 Table.
Mean probability that an author is female for articles in the “Bio” dataset (black dot) or in the “Comp” dataset (open square) for each journal that had at least 1000 authors plotted against the journals’ 2014 impact factor. Journals that had computational biology articles are included in both datasets. Pearson’s correlation was calculated for each dataset independently. Bio: r = -0.246, 95\% CI = [-0.391, -0.088], p = 0.0025. Comp: r = -0.052, 95\% CI = [-0.4458, 0.3592], p = 0.8106.
Mean probability that an author is female for every journal that had at least 1000 authors in our dataset. Grey bars represent journals that have the words “Bioinformatics,” “Computational,” “Computer,” “System(s),” or “-omic(s)” in their title. Vertical line represents the median for female author representation. See also S1 Table.
Mean probability that an author is female for articles in the “Bio” dataset (black dot) or in the “Comp” dataset (open square) for each journal that had at least 1000 authors plotted against the journals’ 2014 impact factor. Journals that had computational biology articles are included in both datasets. Pearson’s correlation was calculated for each dataset independently. Bio: r = -0.246, 95\% CI = [-0.391, -0.088], p = 0.0025. Comp: r = -0.052, 95\% CI = [-0.4458, 0.3592], p = 0.8106.
Figure 3
Mean probability that an author in a given position is female for all preprints in the arXiv quantitative biology (black) or computer science (grey) categories between 2007 and 2014. Error bars represent 95\% confidence intervals.
Mean probability of authors being female in arXiv preprints in a given year. Error bars represent 95\% confidence intervals. Slopes were determined using linear regression. The slope for q-bio is slightly positive (p < 0.5), but the slope for cs is not.
S1 Figure: Gender Inference Validation
Number of primary publications per year indexed under the “Biology” MeSH term.
Number of primary publications per year indexed with “Computational Biology” as a major MeSH term.
Comparison of computational gender inference (black) with known genders (white) for the dataset from Filardo et. al. [12]. Grey represents the known proportion of female authors when excluding names for which the gender could not be computationally inferred. Error bars represent 95\% confidence intervals.
S1 Table: Proportion of Female Authors by Journal
$P_{female}$ for each journal with at least 1000 authors in the bio dataset. Journals identified as primarily computational are shaded grey.
Acknowledgements
Markus Perl for the free use of Gender-API - contact@gender-api.com
Casper Strømgren for the free use of genderize.io info@genderize.io
Giovanni Filardo for sharing data [12]
Johanna Gutlerner, Marshall Thomas, Diane Lam, and other members of the Curriculum Fellows Program (CFP) at Harvard Medical School (HMS) for helpful feedback and discussions
The HMS CFP and Educational Laboratory for resources and mentorship

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{multirow}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{enumitem}
\usepackage{booktabs}

@@

\begin{document}
\maketitle
\section*{Abstract}
While women are generally underrepresented in STEM fields, there are noticeable differences between fields. For instance, the gender ratio in biology is more balanced than in computer science. We were interested in how this difference is reflected in the interdisciplinary field of computational/quantitative biology. To this end, we examined the proportion of female authors in publications from the PubMed and arXiv databases. There are fewer female authors on research papers in computational biology, as compared to biology in general. This is true across authorship position, year, and journal impact factor. A comparison with arXiv shows that quantitative biology papers have a higher ratio of female authors than computer science papers, placing computational biology in between its two parent fields in terms of gender representation. Both in biology and in computational biology, a female last author increases the probability of other authors on the paper being female, pointing to a potential role of female PIs in influencing the gender balance.

\section*{Introduction}
There is ample literature on the underrepresentation of women in STEM fields and the biases contributing to it. Those biases, though often subtle, are pervasive in several ways: they are often held and perpetuated by both men and women, and they are apparent across all aspects of academic and scientific practice. Undergraduate students show bias in favor of men both when rating their peers [1] and their professors [2]. Professors, in turn, are more likely to respond to e-mail from prospective students who are male [3]. They also show gender bias when hiring staff and deciding on a starting salary [4].

@@

\section*{Results and Discussion}
In order to determine if there is a difference in the gender of authors in computational biology compared to biology as a whole, we used data from Pubmed, a database of biology and biomedical publications administered by the US National Library of Medicine. Pubmed uses Medical Subject Heading (MeSH) terms to classify individual papers by subject. The MeSH term “Computational Biology” is a subset of “Biology” and was introduced in 1997, so we restricted our analysis to primary articles published after this date (see Supplementary Figure 1A-B, Materials and Methods).

@@

\begin{table}[]
\centering
\caption{Proportion of Female Authors}
\label{Table 1}
\begin{tabular}{llccc}
\toprule
        &               &       & \multicolumn{2}{c}{95\% CI} \\
\cmidrule(r){4-5}
Dataset & Position      & Mean  & lower        & upper        \\
\midrule
bio     & first         & 0.376 & 0.373        & 0.378        \\
        & second        & 0.379 & 0.376        & 0.381        \\
        & other         & 0.368 & 0.367        & 0.370        \\
        & penultimate   & 0.279 & 0.277        & 0.282        \\
        & last          & 0.245 & 0.243        & 0.247        \\
comp    & first         & 0.316 & 0.312        & 0.320        \\
        & second        & 0.322 & 0.317        & 0.327        \\
        & other         & 0.331 & 0.328        & 0.333        \\
        & penultimate   & 0.236 & 0.231        & 0.241        \\
        & last          & 0.207 & 0.203        & 0.211        \\
\bottomrule
\end{tabular}
\end{table}

One possible explanation for the difference in male and female authorship position might be a difference in role models or mentors. If true, we would expect studies with a female principal investigator to be more likely to attract female collaborators. Conventionally in biology,  the last author on a publication is the principal investigator on the project. Therefore, we looked at two subsets of our data: publications with a female last author ($P_{female}$ > 0.8) and those with a male last author ($P_{female}$ < 0.2). We found that women were substantially more likely to be authors at every other position if the paper had a female last author than if the last author was male (Figure 1C, Table 2). It is possible that female trainees are be more likely to pursue computational biology if they have a mentor that is also female. Since women are less likely to be senior authors, this might reduce the proportion of women overall.

\begin{table}[]
\centering
\caption{Proportion of Female Authors with Female PI}
\label{Table 2}
\begin{tabular}{llcccccc}
\toprule
        &                 & \multicolumn{3}{c}{Male Last Author} & \multicolumn{3}{c}{Female Last Author} \\
\cmidrule(r){3-8}
        &                 &       & \multicolumn{2}{c}{95\% CI}  &        & \multicolumn{2}{c}{95\% CI}   \\
\cmidrule(r){4-5}\cmidrule(r){7-8}
Dataset & Position        & Mean  & lower        & upper         & Mean   & lower        & upper          \\
\midrule
bio     & first           & 0.362 & 0.359        & 0.365          & 0.478 & 0.472        & 0.484          \\
        & second          & 0.359 & 0.357        & 0.362          & 0.460 & 0.454        & 0.466          \\
        & other           & 0.355 & 0.353        & 0.357          & 0.425 & 0.421        & 0.428          \\
        & penultimate     & 0.259 & 0.256        & 0.263          & 0.336 & 0.330        & 0.343          \\
comp    & first           & 0.305 & 0.300        & 0.311          & 0.390 & 0.378        & 0.402          \\
        & second          & 0.306 & 0.300        & 0.312          & 0.379 & 0.366        & 0.392          \\
        & other           & 0.321 & 0.318        & 0.324          & 0.368 & 0.361        & 0.376          \\
        & penultimate     & 0.223 & 0.218        & 0.229          & 0.263 & 0.249        & 0.277          \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[]
\centering
\caption{Proportion of Female Authors in Arxiv}
\label{Table 3}
\begin{tabular}{llccc}
\toprule
        &                  &       & \multicolumn{2}{c}{95\% CI} \\
\cmidrule(r){4-5}
Dataset      & Position    & Mean  & lower        & upper        \\
\midrule
arxivbio     & first       & 0.184 & 0.178        & 0.190        \\
             & second      & 0.210 & 0.200        & 0.219        \\
             & other       & 0.265 & 0.253        & 0.276        \\
             & penultimate & 0.196 & 0.183        & 0.209        \\
             & last        & 0.148 & 0.141        & 0.155        \\
arxivcs      & first       & 0.157 & 0.155        & 0.160        \\
             & second      & 0.175 & 0.172        & 0.179        \\
             & other       & 0.188 & 0.182        & 0.195        \\
             & penultimate & 0.175 & 0.170        & 0.181        \\
             & last        & 0.155 & 0.153        & 0.158        \\
\bottomrule
\end{tabular}
\end{table}

@@

Since biology attracts more women than computer science, we suspect that many women initially decide to study biology and later become interested in computational biology. If this is the case, understanding what factors influence the field of study will provide useful insight when designing interventions to help narrow the gender gap in computer science and computational biology.
\section*{Materials and Methods}

\subsection*{Datasets}

\subsubsection*{Biology publications 1997-2014 (bio)}
This dataset [13] contains all English language publications under the MeSH term "Biology" published between 1997 and 2014, excluding many non-primary sources. Downloaded 12 February, 2016. Search term: ("Biology"[Mesh]) NOT (Review[ptyp] OR Comment[ptyp] OR Editorial[ptyp] OR Letter[ptyp] OR Case Reports[ptyp] OR News[ptyp] OR "Biography" [Publication Type]) AND ("1997/01/01"[PDAT] : "2014/12/31"[PDAT]) AND english[language]

\subsubsection*{Computational biology publications 1997-2014 (comp)}
Same as above [13], except using MeSH term "Computational Biology". Only uses papers where this is is a major term. Date range was selected because this MeSH term was introduced in 1997. Downloaded 12 February, 2016. Search term: ("Computational Biology"[Majr]) NOT (Review[ptyp] OR Comment[ptyp] OR Editorial[ptyp] OR Letter[ptyp] OR Case Reports[ptyp] OR News[ptyp] OR "Biography" [Publication Type]) AND ("1997/01/01"[PDAT] : "2014/12/31"[PDAT]) AND english[language]

\subsubsection*{Medical Papers}
Subset of author and gender data from Filardo et.al [12]. This dataset did not contain author first names or unique publication identifiers. We searched pubmed for the title, author and publication date, and were able to identify 2155/3153 publications to analyze. Publications with no matching search results or with multiple matching search results were excluded.

\subsubsection*{arXiv Quantitative Biology (q-bio)}
This dataset [14] contains all preprints with the label “q-bio” from 2003 (when the section was introduced) to 2014. Downloaded on 10 June, 2016.

\subsubsection*{arXiv CS (cs)}
This dataset [14] contains all preprints with the label “cs” from 2003 to 2014. Downloaded on 10 June, 2016.

\subsection*{Gender Inference}
Genders were determined using Gender-API (http://gender-api.com), which compares first names to a database compiled from government sources as well as from crawling social media profiles and returns a gender probability and a measure of confidence based on the number of times the name appears in the database. The API was queried with the 74,760 unique first names in the dataset (24 May, 2016).

@@

\section*{Figures}

\subsection*{Figure 1}
\begin{enumerate}[label=\Alph*]
\item Mean probability that an author in a given position is female for primary articles indexed in Pubmed with the MeSH term Biology (black) or Computational Biology (grey). The bio dataset is inclusive of papers in the comp dataset. Error bars represent 95\% confidence intervals.
\item Mean probability that an author in a given position is female for primary articles indexed in Pubmed with the MeSH term Biology (black) or Computational Biology (grey). The bio dataset is inclusive of papers in the comp dataset. Error bars represent 95\% confidence intervals.
\item Mean probability that an author is female for publications in a given year. Error bars represent 95\% confidence intervals.
\item Mean probability that the first (F), second (S), penultimate (P) or other (O) author is female for publications where the last author is male ($P_{female}$ < 0.2) or female ($P_{female}$ > 0.8). Papers where the gender of the last author was uncertain or could not be determined were excluded. Error bars represent 95\% confidence intervals.
\end{enumerate}

\subsection*{Figure 2}
\begin{enumerate}[label=\Alph*]
\item Mean probability that an author is female for every journal that had at least 1000 authors in our dataset. Grey bars represent journals that have the words “Bioinformatics,” “Computational,” “Computer,” “System(s),” or “-omic(s)” in their title. Vertical line represents the median for female author representation. See also S1 Table.
Mean probability that an author is female for articles in the “Bio” dataset (black dot) or in the “Comp” dataset (open square) for each journal that had at least 1000 authors plotted against the journals’ 2014 impact factor. Journals that had computational biology articles are included in both datasets. Pearson’s correlation was calculated for each dataset independently. Bio: r = -0.246, 95\% CI = [-0.391, -0.088], p = 0.0025. Comp: r = -0.052, 95\% CI = [-0.4458, 0.3592], p = 0.8106.
\item Mean probability that an author is female for every journal that had at least 1000 authors in our dataset. Grey bars represent journals that have the words “Bioinformatics,” “Computational,” “Computer,” “System(s),” or “-omic(s)” in their title. Vertical line represents the median for female author representation. See also S1 Table.
\item Mean probability that an author is female for articles in the “Bio” dataset (black dot) or in the “Comp” dataset (open square) for each journal that had at least 1000 authors plotted against the journals’ 2014 impact factor. Journals that had computational biology articles are included in both datasets. Pearson’s correlation was calculated for each dataset independently. Bio: r = -0.246, 95\% CI = [-0.391, -0.088], p = 0.0025. Comp: r = -0.052, 95\% CI = [-0.4458, 0.3592], p = 0.8106.
\end{enumerate}

\subsection*{Figure 3}
\begin{enumerate}[label=\Alph*]
\item Mean probability that an author in a given position is female for all preprints in the arXiv quantitative biology (black) or computer science (grey) categories between 2007 and 2014. Error bars represent 95\% confidence intervals.
\item Mean probability of authors being female in arXiv preprints in a given year. Error bars represent 95\% confidence intervals. Slopes were determined using linear regression. The slope for q-bio is slightly positive (p < 0.5), but the slope for cs is not.
\end{enumerate}

\subsection*{S1 Figure: Gender Inference Validation}
\begin{enumerate}[label=\Alph*]
\item Number of primary publications per year indexed under the “Biology” MeSH term.
\item Number of primary publications per year indexed with “Computational Biology” as a major MeSH term.
\item Comparison of computational gender inference (black) with known genders (white) for the dataset from Filardo et. al. [12]. Grey represents the known proportion of female authors when excluding names for which the gender could not be computationally inferred. Error bars represent 95\% confidence intervals.
\end{enumerate}

\subsection*{S1 Table: Proportion of Female Authors by Journal}

$P_{female}$ for each journal with at least 1000 authors in the bio dataset. Journals identified as primarily computational are shaded grey.
Acknowledgements

\section*{Acknowledgements}
The authors would like to thank Markus Perl for the free use of Gender-API - contact@gender-api.com; Casper Strømgren for the free use of genderize.io info@genderize.io; Giovanni Filardo for sharing data [12]; Johanna Gutlerner, Marshall Thomas, Diane Lam, and other members of the Curriculum Fellows Program (CFP) at Harvard Medical School (HMS) for helpful feedback and discussions; and
The HMS CFP and Educational Laboratory for resources and mentorship.
